---
title: "Logbook QC"
author: "Einar Hjörleifsson"
date: "06/04/2020"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


```{r}
library(knitr)
library(vmstools)
library(RColorBrewer)
library(rnaturalearth)
library(sf)
library(tidyverse)
```

```{r}
my_png <- function(file, width, height) {
  png(file, width = width, height = height, 
      res = 800, units = "in", pointsize = 20)
}
```

```{r load-data}
land <- 
  ne_countries("medium", returnclass = "sf") %>% 
  select(admin) %>% 
  st_crop(xmin = -70,
          xmax = 90,
          ymin = 30,
          ymax = 85)
```

```{r}
ICES_LE <- 
  read_csv('delivery/ICES_LE_ISL.csv')
```


```{r logbook-checks}
ICES_LE <- 
  cbind(ICES_LE, vmstools::ICESrectangle2LonLat(ICES_LE$ICES_rectangle, midpoint=TRUE)) %>% 
  as_tibble()
spatBoundLog <- list(xrange = range(ICES_LE$SI_LONG, na.rm=TRUE),
                     yrange = range(ICES_LE$SI_LATI, na.rm=TRUE))
spatCoreLog  <- list(xrange = quantile(ICES_LE$SI_LONG, c(0.025, 0.975), na.rm=TRUE),
                     yrange = quantile(ICES_LE$SI_LATI, c(0.025, 0.975), na.rm=TRUE))
tempBoundLog <- range(ICES_LE$year, na.rm=TRUE)
```

```{r utilities}
source("2019_VMS_ICES-QC/utilities_report.R")
```


## Years & number of records for which data has been submitted:

```{r}
d <- 
  ICES_LE %>% 
  dplyr::count(year) 

d %>% 
  knitr::kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  labs(subtitle = "Logbooks:: Number of annual records",
       x = NULL, y = NULL) +
  scale_x_continuous(breaks = 2000:2030)
```


## Records assigned to invalid Statistical Rectangles

```{r invalid stat-sqrs-le}
if(any(is.na(ICES_LE$SI_LONG))) {
  
  knitr::kable(table(`ICES Rectangle` = ICES_LE$ICES_rectangle[is.na(ICES_LE$SI_LONG)],
                     Year = ICES_LE$year[is.na(ICES_LE$SI_LONG)]), booktabs = TRUE)
  
} else {
  
  x <- "There were no invalid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Distribution of logbook entries by month:

```{r logbook-entries-month, results='asis'}
d <- 
  ICES_LE %>% 
  count(month, year)
d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ month) +
  scale_x_continuous(breaks = 2000:2030) +
  labs(subtitle = "Logbooks: Number of records by months and years",
       x = NULL, y = NULL)
```


## Area extent of data submitted by year:

```{r , results='asis', fig.height = 9}
if (any(!is.na(ICES_LE$SI_LONG))) {
  ICES_LE %>% 
    group_by(year) %>% 
    summarise(across(c("SI_LONG", "SI_LATI"), list(min = min, max = max))) %>% 
    kable()
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Area for which data has been submitted:

```{r , results='asis', fig.height = 9}
if (any(!is.na(ICES_LE$SI_LONG))) {
  p <- 
    ICES_LE %>% 
    select(year, lon = SI_LONG, lat = SI_LATI) %>% 
    distinct() %>% 
    ggplot() +
    theme_minimal() +
    geom_sf(data = land, fill = "grey", colour = "grey") +
    geom_tile(aes(lon, lat), fill = "red3") +
    facet_wrap(~ year) +
    labs(subtitle = "Area for which data has been submitted",
         x = NULL, y = NULL)
  p +
    coord_sf(xlim = spatBoundLog$xrange,
             ylim = spatBoundLog$yrange)
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```



```{r , results='asis',fig.height = 9}
if (any(!is.na(ICES_LE$SI_LONG))) {
  p +
    coord_sf(xlim = spatCoreLog$xrange,
             ylim = spatCoreLog$yrange)
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

\newpage

## Frequency of unique vessel counts by year:

```{r records-by-unique-vessels-le}
if (any(!is.na(ICES_LE$UniqueVessels))) {
  
  d <- 
    ICES_LE %>% 
    count(year, UniqueVessels) 
  d %>% 
    spread(year, n, fill = 0) %>% 
    slice(1:50) %>% 
    kable()
  
  d %>% 
    ggplot(aes(UniqueVessels, n, colour = as.factor(year))) +
    geom_line() +
    scale_y_continuous(trans = "log10") +
    labs(subtitle = "Frequency of unique vessels",
         x = "Unique vessel count",
         y = "Number of records")
  
} else {
  x <- "There were no unique vessel counts provided"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Frequency of vessel length categories by year:

```{r , results='asis'}
d <- 
  ICES_LE %>% 
  rename(vessel_length = vessel_length_category) %>% 
  count(vessel_length, year)
d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ vessel_length, scale = "free_y") +
  labs(subtitle = "Frequency of vessel length categories by year",
       x = NULL, y = "Number of records")
```



## Frequency of gear codes by year:

```{r , results='asis'}
d <- 
  ICES_LE %>% 
  count(year, gear_code)

d %>% 
  spread(year, n, fill = 0) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ gear_code, scale = "free_y") +
  labs(subtitle = "Frequency of gear codes by year",
       x = NULL, y = "Number of records") +
  scale_x_continuous(breaks = 2000:2030)
```

## Number of unique DCF Level 6 codes by year:

```{r , results='asis'}
ICES_LE %>% 
  group_by(year) %>% 
  summarise(n = n_distinct(LE_MET_level6)) %>% 
  kable()
```

## Top 5 DCF Level 6 codes by year:

```{r , results='asis'}
top <- 
  ICES_LE %>% 
  count(LE_MET_level6) %>% 
  arrange(-n) %>% 
  slice(1:12) %>% 
  pull(LE_MET_level6)
d <- 
  ICES_LE %>% 
  filter(LE_MET_level6 %in% top) %>% 
  count(year, LE_MET_level6)
d %>% 
  spread(year, n) %>% 
  kable()
d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ LE_MET_level6, scale = "free_y") +
  scale_x_continuous(breaks = 2000:2030) +
  labs(subtitle = "Top 12 DCF Level 6 codes by year",
       x = NULL, y = "Number of records")
```



## Frequency of VMS enabled category

```{r , results='asis'}
kable(table(ICES_LE$vms_enabled,ICES_LE$vessel_length_category), booktabs = TRUE)
```



## Average fishing days:
```{r , results='asis'}
ICES_LE %>% 
  ggplot(aes(fishing_days)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(subtitle = "Logbooks: Frequency distirbution of fishing days by year",
       x = "Fishing days",
       y = NULL)
```




## Average KW-hours:
```{r , results='asis'}
ICES_LE %>% 
  ggplot(aes(kw_fishing_days)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(subtitle = "Logbooks: Frequency distirbution of kw-fishing days by year",
       x = "Fishing days",
       y = NULL)
```

## Landings by gear by year:
```{r , results='asis'}
d <- 
  ICES_LE %>% 
  group_by(year, gear_code) %>% 
  summarise(landings = round(sum(totweight) / 1e6, 1))
d %>% 
  spread(year, landings) %>% 
  kable()
d %>% 
  ggplot(aes(year, landings)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ gear_code, scale = "free_y") +
  labs(subtitle = "Logbooks: Landings by year and gear",
       x = NULL)
```


## Spatial distribution of effort by year:
```{r , results='asis', fig.height = 9, eval = FALSE}
coordGrd        <- unique(ICES_LE[,c("SI_LONG","SI_LATI","year","ICES_rectangle")])

polRect  <- make_polVMS(coordGrd, resolution = 1)
polRect$year    <- rep(coordGrd$year,each=5)
polRect$ICES_rectangle <- rep(coordGrd$ICES_rectangle, each = 5)

#TIDY ---
dat             <- aggregate(ICES_LE$fishing_days,by=list(ICES_LE$year,ICES_LE$ICES_rectangle),FUN=sum,na.rm=T)

steps               <- ceiling(max(dat$x,na.rm=T)/250)
cutbreaksval        <- unique(c(-1,0,steps*c(1,2.5,5,10,25,50,100,250)))
legval              <- outer(ac(cutbreaksval),ac(cutbreaksval),function(x,y){return(paste(x,"<=",y))})
legval              <- c("0",diag(legval[-1,-c(1,2)]))

palette <- c("white", brewer.pal(length(cutbreaksval)-2,"YlOrRd"))
cols <- palette[cut(dat$x,breaks=cutbreaksval)]
cols                <- cbind(cols,id=1:length(cols),ICES_rectangle=dat$Group.2,year=dat$Group.1)
polRect              <- merge(polRect,cols,by=c("ICES_rectangle","year"))
# ----

spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill = guide_legend(title = "Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = rev(legval)) +
  facet_wrap(~ year, ncol = 2) +
  theme_icesqc(legend.position = "top")
```

```{r}
ICES_LE %>% 
  group_by(year, SI_LONG, SI_LATI) %>% 
  summarise(fd = sum(fishing_days, na.rm = TRUE)) %>% 
  ggplot() +
  geom_sf(data = land, fill = "grey", colour = "grey") +
  geom_tile(aes(SI_LONG, SI_LATI, fill = fd)) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  facet_wrap(~ year) +
  coord_sf(xlim = spatBoundLog$xrange,
           ylim = spatBoundLog$yrange) +
  labs(x = NULL, y = NULL, fill = "fishing\ndays")
```




```{r}
knitr::opts_chunk$set(eval = FALSE)
```



## Spatial difference of effort `r tempBoundLog[1]`:`r (tempBoundLog[2]-1)` vs `r tempBoundLog[2]`
```{r,  fig.height = 9}
base <- with(ICES_LE, 
             aggregate(fishing_days, 
                       by = list(ICES_rectangle = ICES_rectangle, year = year),
                       FUN = sum, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days = x)

# calculate total fishing hours for recent year

recent <- base[base$year == tempBoundLog[2],]

# calculate median of the total fishing hours per square for historical years
base <- with(base[base$year < tempBoundLog[2],], 
             aggregate(fishing_days,
                       by = list(ICES_rectangle = ICES_rectangle), 
                       FUN = median, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days_median = x)

# join
dat2plot <- dplyr::full_join(base,
                             recent[,c("ICES_rectangle","fishing_days")])

# set NAs to zero
dat2plot$fishing_days_median[is.na(dat2plot$fishing_days_median)] <- 0
dat2plot$fishing_days[is.na(dat2plot$fishing_days)] <- 0

# calculate ratio (with exceptions for zeros)
dat2plot$ratio <- 1/with(dat2plot, pmax(fishing_days, 1e-9) / pmax(fishing_days_median, 1e-9))

# add back in lat and long
dat2plot <- cbind(dat2plot,
                  vmstools::ICESrectangle2LonLat(dat2plot$ICES_rectangle, midpoint = TRUE))

# make 'fortified' data frame
polRect <- make_polVMS(dat2plot, resolution = 1)
polRect$ICES_rectangle <- rep(dat2plot$ICES_rectangle, each = 5)

## tidy ---
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")
palette <- brewer.pal(length(cutbreaksval)-1,"RdYlBu")
colgrp <- as.numeric(cut(dat2plot$ratio, breaks = breaks))
cols <- cbind(cols = palette[colgrp], ICES_rectangle = dat2plot$ICES_rectangle)
polRect <- merge(polRect, cols, by=c("ICES_rectangle"))
# ---
# EH should really be spatCoreLog
spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill=guide_legend(title="Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  theme_icesqc(legend.position = "right")
```



## Spatial difference of effort `r tempBoundLog[2]-1` vs `r tempBoundLog[2]`

```{r,  fig.height = 9}
base <- with(ICES_LE, 
             aggregate(fishing_days, 
                       by = list(ICES_rectangle = ICES_rectangle, year = year),
                       FUN = sum, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days = x)

# calculate total fishing hours for recent year
recent <- base[base$year == tempBoundLog[2],]

# previous year
base <- base[base$year == tempBoundLog[2]-1,]
base <- dplyr::rename(base, fishing_days_median = fishing_days)

# join
dat2plot <- dplyr::full_join(base,
                             recent[,c("ICES_rectangle","fishing_days")])

# set NAs to zero
dat2plot$fishing_days_median[is.na(dat2plot$fishing_days_median)] <- 0
dat2plot$fishing_days[is.na(dat2plot$fishing_days)] <- 0

# calculate ratio (with exceptions for zeros)
dat2plot$ratio <- 1/with(dat2plot, pmax(fishing_days, 1e-9) / pmax(fishing_days_median, 1e-9))

# add back in lat and long
dat2plot <- cbind(dat2plot,
                  vmstools::ICESrectangle2LonLat(dat2plot$ICES_rectangle, midpoint = TRUE))

# make 'fortified' data frame
polRect <- make_polVMS(dat2plot, resolution = 1)
polRect$ICES_rectangle <- rep(dat2plot$ICES_rectangle, each = 5)

## tidy ---
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")
palette <- brewer.pal(length(cutbreaksval)-1,"RdYlBu")
colgrp <- as.numeric(cut(dat2plot$ratio, breaks = breaks))
cols <- cbind(cols = palette[colgrp], ICES_rectangle = dat2plot$ICES_rectangle)
polRect <- merge(polRect, cols, by=c("ICES_rectangle"))
# ---

spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill=guide_legend(title="Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  theme_icesqc(legend.position = "right")
```

\newpage


## Relationship fishing days and total weight
```{r , fig.height = 9}
ggplot(ICES_LE[ICES_LE$year == tempBoundLog[2],], 
       aes(x = fishing_days, y = totweight)) +
  geom_point() +
  facet_wrap(~ gear_code, ncol = 3, scale = "free") +
  xlab("Fishing days") + ylab ("Total weight") +
  theme_icesqc()
```

\newpage

## Mean landing per KW fishing day by year:

```{r , results='asis'}
ps <- gear_splits(with(ICES_LE, totweight/kw_fishing_days), data = ICES_LE, "kg landed", gear_groups = 4, func = median)
ps$table
for (p in ps$plots) print(p)
```

\newpage

## Value by gear by year:
```{r , results='asis'}
ps <- gear_splits(with(ICES_LE, totvalue), data = ICES_LE, "EUR landed", gear_groups = 4, func = median)
ps$table
for (p in ps$plots) print(p)
```

\newpage

## Mean value per KW fishing day by year:

```{r , results='asis'}
ps <- gear_splits(with(ICES_LE, totvalue/kw_fishing_days), data = ICES_LE, "EUR/kWh", gear_groups = 4, func = median)
ps$table
for (p in ps$plots) print(p)
```

\newpage

##  Average price:

```{r , results='asis'}
ps <- gear_splits(with(ICES_LE, totvalue/totweight), data = ICES_LE, "Mean price (EUR/kg)", gear_groups = 4, func = median)
cat(ps$table)
for (p in ps$plots) print(p)
```

\newpage

## Comparison of Metier level 6 reporting between logbook and VMS
```{r met6-comparison, results = 'asis', eval = FALSE}

ledat <- with(ICES_LE, table(LE_MET_level6, year))
vedat <- with(ICES_VE, table(LE_MET_level6, year))

dat2tab <- 
  rbind(
    cbind(as.data.frame.table(ledat), data = "LE (records)"),
    cbind(as.data.frame.table(vedat), data = "VE (pings)"))

tab <- with(dat2tab, tapply(Freq, list(LE_MET_level6, data, year), sum))
tab[tab == 0] <- NA

for (i in dimnames(tab)[[3]]) {
  x <- tab[,,i]
  x <- x[apply(!is.na(x), 1, any),]
  cat(kable(cbind(x, year = i), booktabs = TRUE), sep = "\n")
  cat("\n")
}
```

