---
title: "Logbook QC"
author: "Einar Hjörleifsson"
date: "06/04/2020"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
library(knitr)
library(vmstools)
library(RColorBrewer)
library(rnaturalearth)
library(sf)
library(tidyverse)
```


```{r}
land <- 
  ne_countries("medium", returnclass = "sf") %>% 
  select(admin) %>% 
  st_crop(xmin = -70,
          xmax = 90,
          ymin = 30,
          ymax = 85)
```

```{r}
load("2019_VMS_ICES-QC/bootstrap/data/eurPols.Rdata")
polLand <- fortify(eurPols)
```


```{r}
lgs <- 
  read_csv('delivery/ICES_LE_ISL.csv')
```


```{r logbook-checks}
lgs <- 
  cbind(lgs, vmstools::ICESrectangle2LonLat(lgs$ices, midpoint=TRUE)) %>% 
  rename(lon = SI_LONG,
         lat = SI_LATI) %>% 
  as_tibble()
spatBoundLog <- list(xrange = range(lgs$lon, na.rm=TRUE),
                     yrange = range(lgs$lat, na.rm=TRUE))
spatCoreLog  <- list(xrange = quantile(lgs$lon, c(0.025, 0.975), na.rm=TRUE),
                     yrange = quantile(lgs$lat, c(0.025, 0.975), na.rm=TRUE))
tempBoundLog <- range(lgs$year, na.rm=TRUE)
```

```{r utilities}
source("2019_VMS_ICES-QC/utilities_report.R")
```


## Years & number of records for which data has been submitted:

```{r}
d <- 
  lgs %>% 
  dplyr::count(year)

d %>% 
  knitr::kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  labs(subtitle = "Logbooks:: Number of annual records",
       x = NULL, y = NULL) +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2))
```


## Records assigned to invalid Statistical Rectangles

```{r invalid stat-sqrs-le}
if(any(is.na(lgs$lon))) {
  
  knitr::kable(table(`ICES Rectangle` = lgs$ICES_rectangle[is.na(lgs$lon)],
                     Year = lgs$year[is.na(lgs$lon)]), booktabs = TRUE)
  
} else {
  
  x <- "There were no invalid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Distribution of logbook entries by month:

```{r logbook-entries-month, results='asis'}
d <- 
  lgs %>% 
  count(month, year)
d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ month) +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2)) +
  labs(subtitle = "Logbooks: Records by months and years",
       x = NULL, y = NULL)
```


## Area extent of data submitted by year:

```{r , results='asis', fig.height = 9}
if (any(!is.na(lgs$lon))) {
  lgs %>% 
    group_by(year) %>% 
    summarise(across(c("lon", "lat"), list(min = min, max = max))) %>% 
    kable()
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Area for which data has been submitted:

```{r , results='asis', fig.height = 9}
if (any(!is.na(lgs$lon))) {
  p <- 
    lgs %>% 
    select(year, lon = lon, lat = lat) %>% 
    distinct() %>% 
    ggplot() +
    theme_minimal() +
    geom_sf(data = land, fill = "grey", colour = "grey") +
    geom_tile(aes(lon, lat), fill = "red3") +
    facet_wrap(~ year) +
    labs(subtitle = "Area for which data has been submitted",
         x = NULL, y = NULL)
  p +
    coord_sf(xlim = spatBoundLog$xrange,
             ylim = spatBoundLog$yrange)
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```



```{r , results='asis',fig.height = 9}
if (any(!is.na(lgs$lon))) {
  p +
    coord_sf(xlim = spatCoreLog$xrange,
             ylim = spatCoreLog$yrange)
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

\newpage

## Frequency of unique vessel counts by year:

```{r records-by-unique-vessels-le}
if (any(!is.na(lgs$unqv))) {
  
  d <- 
    lgs %>% 
    count(year, unqv) 
  d %>% 
    spread(year, n, fill = 0) %>% 
    slice(1:50) %>% 
    kable()
  
  d %>% 
    ggplot(aes(unqv, n, colour = as.factor(year))) +
    geom_line() +
    scale_y_continuous(trans = "log10") +
    labs(subtitle = "Frequency of unique vessels",
         x = "Unique vessel count",
         y = "Number of records",
         colour = "Year")
  
} else {
  x <- "There were no unique vessel counts provided"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

## Frequency of vessel length categories by year:

```{r , results='asis'}
d <- 
  lgs %>% 
  # NOTE: Make vlclass as factor and order
  rename(vessel_length = vlclass) %>% 
  count(vessel_length, year)
d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ vessel_length, scale = "free_y") +
  labs(subtitle = "Frequency of vessel length categories by year",
       x = NULL, y = "Number of records")
```

## Frequency of gear codes by year:

```{r , results='asis'}
d <- 
  lgs %>% 
  count(year, dcf4)

d %>% 
  spread(year, n, fill = 0) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(subtitle = "Frequency of gear codes by year",
       x = NULL, y = "Number of records") +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2))
```

## Number of unique DCF Level 6 codes by year:

```{r , results='asis'}
lgs %>% 
  group_by(year) %>% 
  summarise(n = n_distinct(dcf6)) %>% 
  kable()
```

## Top 5 DCF Level 6 codes by year:

```{r , results='asis'}
top <- 
  lgs %>% 
  count(dcf6) %>% 
  arrange(-n) %>% 
  slice(1:12) %>% 
  pull(dcf6)
d <- 
  lgs %>% 
  filter(dcf6 %in% top) %>% 
  count(year, dcf6)
d %>% 
  spread(year, n) %>% 
  kable()
d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf6, scale = "free_y") +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2)) +
  labs(subtitle = "Top 12 DCF Level 6 codes by year",
       x = NULL, y = "Number of records")
```



## Frequency of VMS enabled category

```{r , results='asis'}
kable(table(lgs$vms,lgs$vlclass), booktabs = TRUE)
```



## Average fishing days:

NOTE: Should possible invent a better statistical test here, like to end up with whole integer in fishing days.

```{r , results='asis'}
lgs %>% 
  ggplot(aes(fd)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(subtitle = "Logbooks: Frequency distirbution of fishing days",
       x = "Fishing days",
       y = NULL)
```


## Average KW-hours:
```{r , results='asis'}
lgs %>% 
  ggplot(aes(kwfd)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(subtitle = "Logbooks: Frequency distirbution of kw-fishing days",
       x = "Kilowatt fishing days",
       y = NULL)
```

## Landings by gear by year:
```{r , results='asis'}
d <- 
  lgs %>% 
  group_by(year, dcf4) %>% 
  summarise(landings = round(sum(wgt) / 1e6, 1))
d %>% 
  spread(year, landings) %>% 
  kable()
d %>% 
  ggplot(aes(year, landings)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(subtitle = "Logbooks: Landings by year and gear",
       x = NULL)
```


## Spatial distribution of effort by year:

```{r}
p <- 
  lgs %>% 
  group_by(year, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ggplot() +
  geom_sf(data = land, fill = "grey", colour = "grey") +
  geom_tile(aes(lon, lat, fill = fd)) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  facet_wrap(~ year) +
  labs(x = NULL, y = NULL, fill = "fishing\ndays")
p +
  coord_sf(xlim = spatBoundLog$xrange,
           ylim = spatBoundLog$yrange)
```

```{r}
p +
  coord_sf(xlim = spatCoreLog$xrange,
           ylim = spatCoreLog$yrange)
```

## Spatial difference of effort

```{r}
compare_periods <- function(d, yr1 = 2018, yr2 = 2019) {
  
  CUTS <-
    c(-10,
      log(c(1/2, 1/1.75, 1/1.5, 1/1.25, 1/1.05, 1.05, 1.25, 1.5, 1.75, 2)),
      9, 100)
  LABS <-
    c("historic >>", "historic> +100%", "historic> +75%", "historic> +50%", "historic> +25%",
      "+/-5%",
      "recent> +5%","recent> +25%","recent> +50%","recent> +75%", "recent> +100%","recent >>")
  
  d %>%
    filter(year %in% c(yr1, yr2)) %>%
    mutate(period = case_when(year %in% yr1 ~ "t1",
                              year %in% yr2 ~ "t2")) %>%
    group_by(period, lon, lat) %>%
    summarise(fd = sum(fd, na.rm = TRUE)) %>%
    ungroup() %>%
    spread(period, fd, fill = 0) %>%
    mutate(r = case_when(t2 > 0 & t1 == 0 ~ 10,
                         t2 == 0 & t1 > 0 ~ -10,
                         TRUE ~ log(t2 / t1))) %>%
    mutate(rc = cut(r,
                    breaks = CUTS,
                    labels = LABS,
                    include.lowest = TRUE))
}
```


```{r}
lgs %>% 
  compare_periods(yr1 = 2009:2018, yr2 = 2019) %>% 
  ggplot() +
  geom_tile(aes(lon, lat, fill = rc)) +
  geom_path(data = geo::island, aes(lon, lat)) +
  scale_fill_manual(values = c("cyan", colorspace::diverge_hcl(10), "orange")) +
  coord_quickmap() +
  labs(x = NULL, y = NULL,
       subtitle = "Comparison: 2019 vs 2018")
```

```{r}
lgs %>% 
  compare_periods(yr1 = 2018, yr2 = 2019) %>% 
  ggplot() +
  geom_tile(aes(lon, lat, fill = rc)) +
  geom_path(data = geo::island, aes(lon, lat)) +
  scale_fill_manual(values = c("cyan", colorspace::diverge_hcl(10), "orange")) +
  coord_quickmap() +
  labs(x = NULL, y = NULL,
       subtitle = "Comparison: 2019 vs past")
```


```{r}
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")

d <-
  lgs %>% 
  #filter(!dcf4 %in% c("OTM", "PS")) %>% 
  #filter(dcf4 == "OTB") %>% 
  group_by(year, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(period = case_when(year == max(year) ~ "recent",
                            TRUE ~ "historic")) %>% 
  group_by(period, lon, lat) %>% 
  summarise(fd = median(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  spread(period, fd, fill = 0) %>% 
  #mutate(r = 1 / (pmax(recent, 1e-1) / pmax(historic, 1e-1)))
  mutate(r = pmax(recent, 1e-9) / pmax(historic, 1e-9)) %>% 
  mutate(colgrp = cut(r, breaks = breaks))


palette <- brewer.pal(length(breaks)-1,"RdYlBu")
colgrp <- as.numeric(cut(d$r, breaks = breaks))

d %>% 
  ggplot() +
  geom_sf(data = land) +
  geom_tile(aes(lon, lat, fill = colgrp)) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  coord_sf(xlim = range(d$lon), ylim = range(d$lat))

```


```{r}
d <- 
  lgs %>% 
  #filter(!dcf4 %in% c("OTM", "PS")) %>% 
  filter(dcf4 == "OTB") %>% 
  group_by(year, ices, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(period = case_when(year == max(year) ~ "recent",
                            TRUE ~ "historic")) %>% 
  group_by(period, ices, lon, lat) %>% 
  summarise(fd = median(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  spread(period, fd, fill = 0) %>% 
  mutate(r = log( pmax(recent, 1e-4) / pmax(historic, 1e-4))) %>% 
  mutate(r2 = case_when(recent == 0 & historic > 0 ~ -6,
                        recent  > 0 & historic == 0 ~ 6,
                        TRUE ~ log(recent / historic))) 

library(sf)
library(mapview)
ices <- 
  read_sf("/net/www/export/home/ftp/pub/data/shapes/ices_rectangles.gpkg") %>% 
  select(ices = icesname)

d %>% 
  filter(between(r2, -5, 5)) %>% 
  select(ices, r2, historic, recent) %>% 
  mutate(historic = round(historic, 2),
         recent = round(recent, 2)) %>% 
  left_join(ices) %>% 
  st_as_sf() %>% 
  mapview(zcol = "r2", at = seq(-5, 5, 1), alpha.regions = 1, alpha = 0,
          col.regions = colorspace::diverge_hcl(7),
          map.types = "CartoDB.DarkMatter")
```



```{r,  fig.height = 9}
base <- with(lgs, 
             aggregate(fd, 
                       by = list(ICES_rectangle = ices, year = year),
                       FUN = sum, na.rm = TRUE))
base <- dplyr::rename(base, fd = x)

# calculate total fishing hours for recent year

recent <- base[base$year == tempBoundLog[2],]

# calculate median of the total fishing hours per square for historical years
base <- with(base[base$year < tempBoundLog[2],], 
             aggregate(fd,
                       by = list(ICES_rectangle = ICES_rectangle), 
                       FUN = median, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days_median = x)

# join
dat2plot <- dplyr::full_join(base,
                             recent[,c("ICES_rectangle","fd")])

# set NAs to zero
dat2plot$fishing_days_median[is.na(dat2plot$fishing_days_median)] <- 0
dat2plot$fd[is.na(dat2plot$fd)] <- 0

# calculate ratio (with exceptions for zeros)
dat2plot$ratio <- 1/with(dat2plot, pmax(fd, 1e-9) / pmax(fishing_days_median, 1e-9))

# add back in lat and long
dat2plot <- cbind(dat2plot,
                  vmstools::ICESrectangle2LonLat(dat2plot$ICES_rectangle, midpoint = TRUE))

# make 'fortified' data frame
polRect <- make_polVMS(dat2plot, resolution = 1)
polRect$ICES_rectangle <- rep(dat2plot$ICES_rectangle, each = 5)

## tidy ---
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")
palette <- brewer.pal(length(breaks)-1,"RdYlBu")
colgrp <- as.numeric(cut(dat2plot$ratio, breaks = breaks))
cols <- cbind(cols = palette[colgrp], ICES_rectangle = dat2plot$ICES_rectangle)
polRect <- merge(polRect, cols, by=c("ICES_rectangle"))
# ---
# EH should really be spatCoreLog
spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill=guide_legend(title="Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  theme_icesqc(legend.position = "right") +
  coord_quickmap()
```





## Spatial difference of effort `r tempBoundLog[2]-1` vs `r tempBoundLog[2]`

```{r,  fig.height = 9, eval = FALSE}
base <- with(lgs, 
             aggregate(fishing_days, 
                       by = list(ICES_rectangle = ICES_rectangle, year = year),
                       FUN = sum, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days = x)

# calculate total fishing hours for recent year
recent <- base[base$year == tempBoundLog[2],]

# previous year
base <- base[base$year == tempBoundLog[2]-1,]
base <- dplyr::rename(base, fishing_days_median = fishing_days)

# join
dat2plot <- dplyr::full_join(base,
                             recent[,c("ICES_rectangle","fishing_days")])

# set NAs to zero
dat2plot$fishing_days_median[is.na(dat2plot$fishing_days_median)] <- 0
dat2plot$fishing_days[is.na(dat2plot$fishing_days)] <- 0

# calculate ratio (with exceptions for zeros)
dat2plot$ratio <- 1/with(dat2plot, pmax(fishing_days, 1e-9) / pmax(fishing_days_median, 1e-9))

# add back in lat and long
dat2plot <- cbind(dat2plot,
                  vmstools::ICESrectangle2LonLat(dat2plot$ICES_rectangle, midpoint = TRUE))

# make 'fortified' data frame
polRect <- make_polVMS(dat2plot, resolution = 1)
polRect$ICES_rectangle <- rep(dat2plot$ICES_rectangle, each = 5)

## tidy ---
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")
palette <- brewer.pal(length(cutbreaksval)-1,"RdYlBu")
colgrp <- as.numeric(cut(dat2plot$ratio, breaks = breaks))
cols <- cbind(cols = palette[colgrp], ICES_rectangle = dat2plot$ICES_rectangle)
polRect <- merge(polRect, cols, by=c("ICES_rectangle"))
# ---

spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill=guide_legend(title="Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  theme_icesqc(legend.position = "right")
```


## Relationship fishing days and total weight
```{r , fig.height = 9}
lgs %>% 
  filter(year == max(year)) %>% 
  ggplot() +
  geom_point(aes(fd, wgt / 1e3)) +
  facet_wrap(~ dcf4, scale = "free") +
  labs(x = "Fishing days",
       y = "Total weight [tonnes]",
       subtitle = "Relationship of fishing days and total weight")
```


## Mean landing per KW fishing day by year:

```{r , results='asis'}
lgs %>% 
  mutate(wgt.pkwd = wgt / kwfd) %>% 
  ggplot(aes(year, wgt.pkwd)) +
  stat_summary(fun.data = "mean_cl_boot", size = 1) +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(x = NULL, y = "[kg]",
       subtitle = "Mean landing per KW fishing day",
       caption = "Bootstrap mean and confidence interval")
```

\newpage

## Value by gear by year:

NOTE: This should be similar to landings by gear

```{r , results='asis'}

```


## Mean value per KW fishing day by year:

```{r , results='asis'}
lgs %>% 
  mutate(value.pkwd = wgt / kwfd) %>% 
  ggplot(aes(year, value.pkwd)) +
  stat_summary(fun.data = "mean_cl_boot", size = 1) +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(x = NULL, y = "[kg]",
       subtitle = "Mean value per KW fishing day",
       caption = "Bootstrap mean and confidence interval")
```


##  Average price:

```{r , results='asis', eval = FALSE}
ps <- gear_splits(with(lgs, value/wgt), data = lgs, "Mean price (EUR/kg)", gear_groups = 4, func = median)
cat(ps$table)
for (p in ps$plots) print(p)
```

\newpage

## Comparison of Metier level 6 reporting between logbook and VMS
```{r met6-comparison, results = 'asis', eval = FALSE}

ledat <- with(lgs, table(LE_MET_level6, year))
vedat <- with(ICES_VE, table(LE_MET_level6, year))

dat2tab <- 
  rbind(
    cbind(as.data.frame.table(ledat), data = "LE (records)"),
    cbind(as.data.frame.table(vedat), data = "VE (pings)"))

tab <- with(dat2tab, tapply(Freq, list(LE_MET_level6, data, year), sum))
tab[tab == 0] <- NA

for (i in dimnames(tab)[[3]]) {
  x <- tab[,,i]
  x <- x[apply(!is.na(x), 1, any),]
  cat(kable(cbind(x, year = i), booktabs = TRUE), sep = "\n")
  cat("\n")
}
```

