---
title: "Logbook QC"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    code_folding: hide
    toc: yes
    toc_float: yes
    toc_depth: 3
params:
  logbook_filename: must_supply
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(knitr)
#library(kableExtra)
library(vmstools)
#library(RColorBrewer)
library(rnaturalearth)
library(sf)
library(mapview)
library(tidyverse)
```


```{r}
land <- 
  rnaturalearth::ne_countries("medium", returnclass = "sf") %>% 
  select(admin) %>% 
  st_crop(xmin = -70,
          xmax = 90,
          ymin = 30,
          ymax = 85)
```

```{r}
lgs <- 
  read_csv(params$logbook_filename)

  # some renaming
  lgs <-
    lgs %>%
    rename(
      ices = ICES_rectangle,
      unqv = UniqueVessels,
      vlclass = vessel_length_category,
      dcf4 = gear_code,
      dcf6 = LE_MET_level6,
      fd = fishing_days,
      kwfd = kw_fishing_days,
      wgt = totweight,
      value = totvalue,
      vms = vms_enabled
    )
```


## Preamble


**NOTE to oneself**: The fundamental record (a single row) is statistics by year, month, statistical rectangel, dcf6 and vessel category.

**TODO**: Complete spatial comparision between year

A peek at the data:

```{r}
glimpse(lgs)
```

```{r logbook-checks}
lgs <- 
  cbind(lgs, vmstools::ICESrectangle2LonLat(lgs$ices, midpoint=TRUE)) %>% 
  rename(lon = SI_LONG,
         lat = SI_LATI) %>% 
  as_tibble()
spatBoundLog <- list(xrange = range(lgs$lon, na.rm=TRUE),
                     yrange = range(lgs$lat, na.rm=TRUE))
spatCoreLog  <- list(xrange = quantile(lgs$lon, c(0.025, 0.975), na.rm=TRUE),
                     yrange = quantile(lgs$lat, c(0.025, 0.975), na.rm=TRUE))
tempBoundLog <- range(lgs$year, na.rm=TRUE)
```

## Time

#### **Records by year**

```{r, fig.height = 3}
d <- 
  lgs %>% 
  dplyr::count(year)

d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot() +
  theme_gray(base_size = 14) +
  geom_pointrange(aes(x = year, y = n,
                      ymin = 0.95 * min(d$n), ymax = n),
                  size = 1.5) +
  labs(x = NULL, y = "Number of records") +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2))
```

#### **Records by year and month**

```{r}
d <- 
  lgs %>% 
  count(month, year)
d %>% 
  spread(month, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  theme_grey(base_size = 14) +
  geom_point() +
  geom_line() +
  facet_wrap(~ month) +
  scale_x_continuous(breaks = seq(2000, 2030, by = 4)) +
  labs(caption = "Records by year and month",
       x = NULL, y = NULL)
```

## Vessels

#### **Frequency of unique vessel counts by year**

```{r, fig.height = 4}
if (any(!is.na(lgs$unqv))) {
  
  d <- 
    lgs %>% 
    count(year, unqv) 
  d %>% 
    spread(year, n, fill = 0) %>% 
    slice(1:50) %>% 
    kable()
  
  d %>% 
    ggplot(aes(unqv, n, colour = as.factor(year))) +
    geom_line() +
    scale_y_continuous(trans = "log10") +
    scale_x_continuous(trans = "log10") +
    labs(caption = "Frequency of unique vessels",
         x = "Unique vessel count",
         y = "Number of records",
         colour = "Year") +
    theme(legend.position = c(0.95, 0.65))
  
} else {
  x <- "There were no unique vessel counts provided"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

#### **Frequency of vessel length categories by year**

```{r}
d <- 
  lgs %>% 
  # NOTE: Make vlclass as factor and order
  rename(vessel_length = vlclass) %>% 
  count(vessel_length, year)
d %>% 
  spread(year, n) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ vessel_length, scale = "free_y") +
  labs(caption = "Frequency of vessel length categories by year",
       x = NULL, y = "Number of records")
```

## DCFs

### DCF4

#### **Number of unique gear (DCF4) codes by year**

```{r}
lgs %>% 
  group_by(year) %>% 
  summarise(n = n_distinct(dcf4)) %>% 
  spread(year, n) %>% 
  kable()
```

#### **Number of records by gear code (DCF4)**

```{r}
d <- 
  lgs %>% 
  count(year, dcf4)

d %>% 
  spread(year, n, fill = 0) %>% 
  kable()

d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf4, scale = "free_y", ncol = 3) +
  labs(caption = "Records by gear code (DCF4) by year",
       x = NULL, y = "Number of records") +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2))
```

### DCF6

#### **Number of unique DCF Level 6 codes by year**

```{r}
lgs %>% 
  group_by(year) %>% 
  summarise(n = n_distinct(dcf6)) %>% 
  spread(year, n) %>% 
  kable()
```

#### **Top 12 DCF Level 6 codes by year**

```{r , results='asis'}
top <- 
  lgs %>% 
  count(dcf6) %>% 
  arrange(-n) %>% 
  slice(1:12) %>% 
  pull(dcf6)
d <- 
  lgs %>% 
  filter(dcf6 %in% top) %>% 
  count(year, dcf6)
d %>% 
  spread(year, n) %>% 
  kable()
d %>% 
  ggplot(aes(year, n)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf6, scale = "free_y", ncol = 3) +
  scale_x_continuous(breaks = seq(2000, 2030, by = 2)) +
  labs(caption = "Top 12 DCF Level 6 codes by year",
       x = NULL, y = "Number of records")
```


## Statistics

#### **Frequency distribution of fishing days**

```{r , results='asis'}
lgs %>% 
  ggplot(aes(fd)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(x = "Fishing days",
       y = NULL)
```

#### **Frequency distirbution of kw-fishing days**

```{r , results='asis'}
lgs %>% 
  ggplot(aes(kwfd)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10") +
  facet_wrap(~ year) +
  labs(x = "Kilowatt fishing days",
       y = NULL)
```

#### **Landings [tonnes] by gear by year**

```{r}
d <- 
  lgs %>% 
  group_by(year, dcf4) %>% 
  summarise(landings = round(sum(wgt) / 1e3, 1))
d %>% 
  spread(year, landings) %>% 
  kable()
d %>% 
  ggplot(aes(year, landings)) +
  geom_point() +
  geom_line() +
  facet_wrap(~ dcf4, scale = "free_y", ncol = 3) +
  labs(caption = "Landings by year and gear",
       x = NULL,
       y = "[tonnes]")
```

#### **Mean landing per fishing day by year**

```{r}
lgs %>% 
  mutate(wgt.pkfd = (wgt / 1e3) / fd) %>% 
  ggplot(aes(year, wgt.pkfd)) +
  stat_summary(fun.data = "mean_cl_boot", size = 1) +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(x = NULL, y = "[tonnes / day]",
       caption = "Bootstrap mean and confidence interval")
```

#### **Relationship fishing days and total weight**

```{r}
lgs %>% 
  filter(year == max(year)) %>% 
  ggplot() +
  geom_point(aes(fd, wgt / 1e3)) +
  facet_wrap(~ dcf4, scale = "free") +
  labs(x = "Fishing days",
       y = "Total weight [tonnes]")
```

#### **Mean landing per KW fishing day by year**

```{r}
lgs %>% 
  mutate(wgt.pkwd = (wgt / 1e3) / kwfd) %>% 
  ggplot(aes(year, wgt.pkwd)) +
  stat_summary(fun.data = "mean_cl_boot", size = 1) +
  facet_wrap(~ dcf4, scale = "free_y") +
  labs(x = NULL, y = "[kg]",
       caption = "Bootstrap mean and confidence interval")
```

#### **Value by gear by year**

NOTE: This should be similar to landings by gear

```{r}

```


#### **Mean value per KW fishing day by year**

```{r}
if(any(!is.na(lgs$value))) {
  lgs %>% 
    mutate(value.pkwd = value / kwfd) %>% 
    ggplot(aes(year, value.pkwd)) +
    stat_summary(fun.data = "mean_cl_boot", size = 1) +
    facet_wrap(~ dcf4, scale = "free_y") +
    labs(x = NULL, y = "[kg]",
         caption = "Bootstrap mean and confidence interval")
}
```


####  **Average price**

```{r eval = FALSE}
ps <- gear_splits(with(lgs, value/wgt), data = lgs, "Mean price (EUR/kg)", gear_groups = 4, func = median)
cat(ps$table)
for (p in ps$plots) print(p)
```

## Space

#### **Invalid statistical rectangles**

```{r invalid stat-sqrs-le}
if(any(is.na(lgs$lon))) {
  
  knitr::kable(table(`ICES Rectangle` = lgs$ices[is.na(lgs$lon)],
                     Year = lgs$year[is.na(lgs$lon)]), booktabs = TRUE)
  
} else {
  
  x <- "There were no invalid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

#### **Spatial extent of data submitted by year**

```{r, eval=FALSE}
lgs %>% 
  group_by(year) %>% 
  summarise(across(c("lon", "lat"), list(min = min, max = max))) %>% 
  kable()
```

#### **Overall range**

```{r}
if (any(!is.na(lgs$lon))) {
  p1 <- 
    lgs %>% 
    select(year, lon = lon, lat = lat) %>% 
    distinct() %>% 
    ggplot() +
    theme_minimal(base_size = 14) +
    geom_sf(data = land, fill = "grey", colour = "grey") +
    geom_tile(aes(lon, lat), fill = "red3") +
    facet_wrap(~ year)
  p1 +
    coord_sf(xlim = spatBoundLog$xrange,
             ylim = spatBoundLog$yrange) +
    labs(x = NULL, y = NULL)
} else {
  x <- "There were no valid Statistical Rectangles reported"
  attr(x, "format") <- "markdown"
  attr(x, "class") <- "knit_asis"
  x
}
```

#### **Effort - overall range**

```{r}
p2 <- 
  lgs %>% 
  group_by(year, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ggplot() +
  geom_sf(data = land, fill = "grey", colour = "grey") +
  geom_tile(aes(lon, lat, fill = fd)) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  facet_wrap(~ year) +
  labs(x = NULL, y = NULL, fill = "fishing\ndays")
p2 +
  coord_sf(xlim = spatBoundLog$xrange,
           ylim = spatBoundLog$yrange)
```

#### **Core range**

```{r}
p1 +
  coord_sf(xlim = spatCoreLog$xrange,
           ylim = spatCoreLog$yrange) +
  labs(x = NULL, y = NULL)
```

### **Effort - core range**

```{r}
p2 +
  coord_sf(xlim = spatCoreLog$xrange,
           ylim = spatCoreLog$yrange)
```

#### Spatial difference of effort

**INCOMPLETE**

```{r}
compare_periods <- function(d, yr1 = 2018, yr2 = 2019) {
  
  CUTS <-
    c(-10,
      log(c(1/2, 1/1.75, 1/1.5, 1/1.25, 1/1.05, 1.05, 1.25, 1.5, 1.75, 2)),
      9, 100)
  LABS <-
    c("historic >>", "historic> +100%", "historic> +75%", "historic> +50%", "historic> +25%",
      "+/-5%",
      "recent> +5%","recent> +25%","recent> +50%","recent> +75%", "recent> +100%","recent >>")
  
  d %>%
    filter(year %in% c(yr1, yr2)) %>%
    mutate(period = case_when(year %in% yr1 ~ "t1",
                              year %in% yr2 ~ "t2")) %>%
    group_by(period, lon, lat) %>%
    summarise(fd = sum(fd, na.rm = TRUE)) %>%
    ungroup() %>%
    spread(period, fd, fill = 0) %>%
    mutate(r = case_when(t2 > 0 & t1 == 0 ~ 10,
                         t2 == 0 & t1 > 0 ~ -10,
                         TRUE ~ log(t2 / t1))) %>%
    mutate(rc = cut(r,
                    breaks = CUTS,
                    labels = LABS,
                    include.lowest = TRUE))
}
```

**INCOMPLETE**

```{r}
lgs %>% 
  compare_periods(yr1 = 2009:2018, yr2 = 2019) %>% 
  ggplot() +
  geom_tile(aes(lon, lat, fill = rc)) +
  geom_path(data = geo::island, aes(lon, lat)) +
  scale_fill_manual(values = c("cyan", colorspace::diverge_hcl(10), "orange")) +
  coord_quickmap() +
  labs(x = NULL, y = NULL,
       title = "Comparison: 2019 vs 2018")
```

**INCOMPLETE**

```{r}
lgs %>% 
  compare_periods(yr1 = 2018, yr2 = 2019) %>% 
  ggplot() +
  geom_tile(aes(lon, lat, fill = rc)) +
  geom_path(data = geo::island, aes(lon, lat)) +
  scale_fill_manual(values = c("cyan", colorspace::diverge_hcl(10), "orange")) +
  coord_quickmap() +
  labs(x = NULL, y = NULL,
       title = "Comparison: 2019 vs past")
```

**INCOMPLETE**

```{r}
library(RColorBrewer)
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")

d <-
  lgs %>% 
  #filter(!dcf4 %in% c("OTM", "PS")) %>% 
  #filter(dcf4 == "OTB") %>% 
  group_by(year, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(period = case_when(year == max(year) ~ "recent",
                            TRUE ~ "historic")) %>% 
  group_by(period, lon, lat) %>% 
  summarise(fd = median(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  spread(period, fd, fill = 0) %>% 
  #mutate(r = 1 / (pmax(recent, 1e-1) / pmax(historic, 1e-1)))
  mutate(r = pmax(recent, 1e-9) / pmax(historic, 1e-9)) %>% 
  mutate(colgrp = cut(r, breaks = breaks))


palette <- brewer.pal(length(breaks)-1,"RdYlBu")
colgrp <- as.numeric(cut(d$r, breaks = breaks))

d %>% 
  ggplot() +
  geom_sf(data = land) +
  geom_tile(aes(lon, lat, fill = colgrp)) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  coord_sf(xlim = range(d$lon), ylim = range(d$lat))

```

**INCOMPLETE**

```{r, eval = FALSE}
d <- 
  lgs %>% 
  #filter(!dcf4 %in% c("OTM", "PS")) %>% 
  filter(dcf4 == "OTB") %>% 
  group_by(year, ices, lon, lat) %>% 
  summarise(fd = sum(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(period = case_when(year == max(year) ~ "recent",
                            TRUE ~ "historic")) %>% 
  group_by(period, ices, lon, lat) %>% 
  summarise(fd = median(fd, na.rm = TRUE)) %>% 
  ungroup() %>% 
  spread(period, fd, fill = 0) %>% 
  mutate(r = log( pmax(recent, 1e-4) / pmax(historic, 1e-4))) %>% 
  mutate(r2 = case_when(recent == 0 & historic > 0 ~ -6,
                        recent  > 0 & historic == 0 ~ 6,
                        TRUE ~ log(recent / historic))) 


ices <- 
  read_sf("/net/www/export/home/ftp/pub/data/shapes/ices_rectangles.gpkg") %>% 
  select(ices = icesname)

d %>% 
  filter(between(r2, -5, 5)) %>% 
  select(ices, r2, historic, recent) %>% 
  mutate(historic = round(historic, 2),
         recent = round(recent, 2)) %>% 
  left_join(ices) %>% 
  st_as_sf() %>% 
  mapview(zcol = "r2", at = seq(-5, 5, 1), alpha.regions = 1, alpha = 0,
          col.regions = colorspace::diverge_hcl(7),
          map.types = "CartoDB.DarkMatter")
```



#### Spatial difference of effort `r tempBoundLog[2]-1` vs `r tempBoundLog[2]`

```{r,  fig.height = 9, eval = FALSE}
base <- with(lgs, 
             aggregate(fishing_days, 
                       by = list(ICES_rectangle = ICES_rectangle, year = year),
                       FUN = sum, na.rm = TRUE))
base <- dplyr::rename(base, fishing_days = x)

# calculate total fishing hours for recent year
recent <- base[base$year == tempBoundLog[2],]

# previous year
base <- base[base$year == tempBoundLog[2]-1,]
base <- dplyr::rename(base, fishing_days_median = fishing_days)

# join
dat2plot <- dplyr::full_join(base,
                             recent[,c("ICES_rectangle","fishing_days")])

# set NAs to zero
dat2plot$fishing_days_median[is.na(dat2plot$fishing_days_median)] <- 0
dat2plot$fishing_days[is.na(dat2plot$fishing_days)] <- 0

# calculate ratio (with exceptions for zeros)
dat2plot$ratio <- 1/with(dat2plot, pmax(fishing_days, 1e-9) / pmax(fishing_days_median, 1e-9))

# add back in lat and long
dat2plot <- cbind(dat2plot,
                  vmstools::ICESrectangle2LonLat(dat2plot$ICES_rectangle, midpoint = TRUE))

# make 'fortified' data frame
polRect <- make_polVMS(dat2plot, resolution = 1)
polRect$ICES_rectangle <- rep(dat2plot$ICES_rectangle, each = 5)

## tidy ---
breaks <- rev(c(1e-10,0.5,2/3,0.8,0.952381,1,1.05,1.25,1.5,2,1e10))
legval <- c("historic >>","historic> +100%","historic> +50%","historic> +25%",
            "+/-5%",
            "recent> +5%","recent> +25%","recent> +50%","recent> +100%","recent >>")
palette <- brewer.pal(length(cutbreaksval)-1,"RdYlBu")
colgrp <- as.numeric(cut(dat2plot$ratio, breaks = breaks))
cols <- cbind(cols = palette[colgrp], ICES_rectangle = dat2plot$ICES_rectangle)
polRect <- merge(polRect, cols, by=c("ICES_rectangle"))
# ---

spatialplot(polRect, xyrange = spatBoundLog) +
  guides(fill=guide_legend(title="Days@Sea")) +
  scale_fill_manual(values = rev(palette), labels = legval) +
  theme_icesqc(legend.position = "right")
```



## VMS

```{r}
lgs %>% 
  count(vlclass, vms) %>% 
  spread(vlclass, n) %>% 
  kable(caption = "Number of records by availability of VMS")
```


#### Comparison of Metier level 6 reporting between logbook and VMS

```{r met6-comparison, eval = FALSE}

ledat <- with(lgs, table(LE_MET_level6, year))
vedat <- with(ICES_VE, table(LE_MET_level6, year))

dat2tab <- 
  rbind(
    cbind(as.data.frame.table(ledat), data = "LE (records)"),
    cbind(as.data.frame.table(vedat), data = "VE (pings)"))

tab <- with(dat2tab, tapply(Freq, list(LE_MET_level6, data, year), sum))
tab[tab == 0] <- NA

for (i in dimnames(tab)[[3]]) {
  x <- tab[,,i]
  x <- x[apply(!is.na(x), 1, any),]
  cat(kable(cbind(x, year = i), booktabs = TRUE), sep = "\n")
  cat("\n")
}
```

